---
title: "R best coding practices"
author: "PLR"
date: "`r format(Sys.time(),'%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Best coding practices

As a closing session to my amazing secondment experience with the NHS Midlands team, I would like to collate a set of standardised procedures that have helped me to write better R code and to hand over my projects to R developers in the team.

These are a set of topics I consider to be important for both beginners and experienced R analysts

- 1. Setup your work based on projects
- 2. Organize your input and outputs around relative paths to your project
- 3. Import Excel files into R using Janitor package  
- 4. Create functions to automate reports done in Excel
- 5. Always check how your scripts use your computer resources by using PROVFIS  package
- 6. Use Git or Github to apply version control on any ongoing script created by several developers 
- 7. Create your own documentation in Markdown and get used to share and review it among all team members
- 8. Run continuous code optimisation: Code once, re-use in the form of functions or even build your own packages for adhoc reports. Then make this packages available for the whole NHS R community.
- 9. Share your code as much as possible from your Github or Git repositories

### 1. Setup your work based on projects

An R project enables your work to be bundled in a portable, self-contained folder. Within the project, all the relevant scripts, data files, figures/outputs, and history are stored in sub-folders and importantly - the working directory is the project’s root folder.

Read this website below to learn more about project oriented workflows:

**Project oriented workflows** <https://rstats.wtf/project-oriented-workflow.html>

#### 1.1 Use functions to setup project folder structures

One of the first steps in producing a reproducible workflow in a team, is to create the same set of folder across project. This is a template useful to organise any type of project, that can be tailored depending on the specific aim of the project. 

```{r Setup your folder structure, echo=TRUE}
library(here)

here::i_am("00 Project folder setup.R")

# 01 Create folder structure making use of HERE package 
project_folder_setup <-function(){
  
  if(!dir.exists("Data")){dir.create(here::here("Data"))}
  if(!dir.exists("Markdown")){dir.create(here::here("Markdown"))}
  if(!dir.exists("Functions")){dir.create(here::here("Functions"))}
  if(!dir.exists("Test")){dir.create(here::here("Test"))}
  if(!dir.exists("Report")){dir.create(here::here("Report"))}
  if(!dir.exists("Model")){dir.create(here::here("Model"))}
  
}
# Execute function
project_folder_setup()
```

### 2. Organize your input and outputs around relative paths 

Always start working on a project thinking about relative paths. You want to use relative paths to run your script in any machine by any developer. You will unlearn the use of setwd() and getwd() functions, as they tend to make people use absolute paths. Instead I recommend you to learn about the multiple applications of the here() package.

Now that we see how useful here() function can be let's use it to write in and wrote out  .csv data

```{r Create some dummy data, echo=TRUE}
# Create a dummy data set

Indic <-c(replicate(2,"Knee Replacement"),
             replicate(2,"Hip Replacement"),
             replicate(2,"Falls and fractures"))
Org <- c(replicate(2,"NHS Northumberland CCG"),
            replicate(2,"NHS Northumberland CCG"),
            replicate(2,"NHS Northumberland CCG"))
Value <-c(rnorm(3))
# Create for each Indicators two rows one for Org value and another for national value
Scope <- c(replicate(6,c("Org","national")))

Mydata <-cbind.data.frame(Indic,Org,Value,Scope)
Mydata
```
We are going to write the newly created Mydata data set into our Data folder we have created when setting up our folder structure. 

```{r Write in and out files using here package, echo=TRUE}
# First we load here package
library(here)

# I assume that you work always using PROJECTS, so the here library writes file path relative to your project folder
# a) Save output in a sub_folder of your main project folder directory
# Saving .csv file in a directory sub_folder called "data"
# /data
write.csv(Mydata,here("data","Mydata_transformed.csv"), row.names = TRUE)

# b) Save output in a deeper sub_folder of your main project folder directory
# /data/sample/

write.csv(Mydata,here("data","sample","Mydata_transformed.csv"), row.names = TRUE)

# In the same way, we can read in the file we have exported into R using the same logic

# Read in (.csv) file using here()
read_in_test  <-read.table(here("Data", "Mydata.csv"),
                           header =TRUE, sep =',',stringsAsFactors =TRUE)
read_in_test
```

These are some useful online resources to learn about the here package
* **Here library** <https://here.r-lib.org/>.

#### 2.1 Some useful applications of Here library 

* a. Get your reference directory 

```{r}
here::i_am("01 Best coding practises.Rmd")
library(here)
```
* b. Write out paths to input and output files

As we have just seen, we can use here() library to write relative paths when bringin in data into R or writting out some results

### 3. Import Excel files into R using readxl and Janitor packages  

When importing data into R from Excel some NHS data sets have long and winded columns names. It is important to tidy them up before starting any data processing. This is why we import Excel files using **readxl** and **janitor**  packages to obtain consistent variables names from Excel. The **readxl** package allows us to import Excel file, controlling with parts and sections of the Excel file we ingest into R and **janitor** library provides a consistent naming convention for columns. This will ease early data wrangling steps and also that we work across project with same set of variable names.  

```{r Import Excel files_when we do NOT specify any paramter, echo=TRUE}
library(readxl)
library(here)
library(dplyr)
library(janitor)

Path <- here()
Path

# Original Excel file "WTL-Data-202021-(20220526)-v1.13.xlsx", saved in data folder
# in our R Project
# Step 01 02: read_excel() function from readxl package
# Monthly Hospital Activity Data
# https://www.england.nhs.uk/statistics/statistical-work-areas/hospital-activity/monthly-hospital-activity/mar-data/
# Source: NHS England and NHS Improvement: monthly RTT data collection
# April 2007 to June 2022

# Consultant-led Referral to Treatment Waiting times Data 2022-23
Referrals_data <- read_excel(here("data", "RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx"), sheet = 1)
Referrals_data
```

As we can see the data structure of the input data set is not clear and column names are not arranged as our original Excel input file.

#### 3.1 Choose which sheet to import

We use sheet command to choose which of the total number of sheet we want to import. In this instance it is just sheet one as we have one tab

This **sheet** command goes after a comma next to the full path section to the Excel file

```{r Import excel read_excel sheet parameter, echo=TRUE}
library(here)
Myexcel <- read_excel(here("data","RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx"), sheet =1 )
```

We still have an issue with *empty* rows being displayed in our R input file. We need to fix this in the next section

** List first number of sheet using excel_sheets()

Fist we need to decide which of the several sheets in the Excel file we are interested in importing into R
It is good practise to list first the sheet names to decide which one we pick for our analysis

```{r list sheet names, echo=TRUE}
excel_sheets('./data/RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx')
```

In this instance, we can see that we have got just one Sheet in our Excel file 

#### 3.2 Omit empty rows above the target table

It is common that these Excel files include  several rows explanations about the data displayed on the table. When importing this table into R we want to skip those lines and start reading data into R from the Table itself. 
In our example using  **RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx** file we see that our data starts on row 10, so we need to specify it to the read_excel function

```{r Import excel read_excel skip parameter, echo=TRUE}
library(here)
Myexcel <- read_excel(here("data","RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx"), sheet =1,skip=9 )
```

Now we can see how the data has been imported by the variable names displayed as columns doesn't show a consistent naming convention. This is why we need to use always   **clean_names()** function from janitor library in combination with **read_excel()** function

```{r Import clean names from Excel using Janitor, echo=TRUE}
# Step 02 02: import  data by using clean_names() function from janitor() package 
library(here)
library(janitor)
Myexcel_clean <- read_excel(here("data","RTT-Overview-Timeseries-Jun22-XLS-77K-68395.xlsx"), sheet =1,skip=9 ) %>% 
                 clean_names()
Myexcel_clean                 
```
Now we have some type of structure that will allows us to carry on our Exploratory analysis, and the variables have consistent names that we can start to rename in a sistematic way. But at this point we have done some pre-processing of the Excel input file that will help us in later stages of our analysis. 

This is the set of variables we have now in our R session
```{r Varaible names after read_excel and clean_names, echo=TRUE}
names(Myexcel_clean)
```

